A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-02-10 article-information-2019.tex
submitted, revised version

% Authors, for the paper (add full first names)
\Author{Navdeep Gill $^{1, \dagger}$, Patrick Hall $^{1,3,\dagger,*}$, Kim Montgomery $^{1,\dagger}$, and Nicholas Schmidt $^{2, \dagger}$}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBMs) and explainable neural networks (XNNs), a deep learning architecture well-suited for structured data, are assessed on simulated data and publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and with global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents, adverse impact ratio (AIR), marginal effect (ME), and standardized mean difference (SMD), and also with straightforward group fairness measures. By combining interpretable models, post-hoc explanations, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/gx/en/issues/data-and-analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}, and ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of ML opaqueness, discrimination, privacy harms, and security vulnerabilities have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, mostly in the form interpretable models, post-hoc explanation, and DI and discrimination testing, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve Supervisory and Regulatory (SR) Letter 11-7, and the European Union (EU) General Data Protection Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

The same can be said of the selected explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. Presented explanation techniques include PD and ICE (see Subsection \ref{ssec:pd_ice}) and Shapley values (see Subsection \ref{ssec:shap}) \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods (see Subsection \ref{ssec:di}) include AIR, ME, and SMD \cite{feldman2015certifying}, \cite{cohen1988statistical}, \cite{cohen1992power}.\footnote{See: Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix measures are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available on GitHub (see Subsection \ref{ssec:soft}).

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e., GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}. XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some instance , $\mathbf{x}$, is defined as:

@@

Because many current technical discussions of fairness in ML appear inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in regulated, high-stakes employment and financial decisions. The discussed measures are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination in US legal and regulatory settings and Appendix \ref{a_sec:prac_v_stat} for remarks on practical vs. statistical significance for discrimination measures. One such common measure of DI used in US litigation and regulatory settings is ME. ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

\noindent where $\hat{\mathbf{y}}$ are the model decisions, $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), $p$ indicates a protected group, and $\text{Pr}(\cdot)$ is the operator for conditional probability.  ME is a favored DI measure used by the the US Consumer Financial Protection Bureau (CFPB), the primary agency charged with regulating fair lending laws at the largest US lending institutions and for various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important DI measure is AIR, more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

\noindent AIR is equal to the ratio of the proportion of the protected class that receives a favorable outcome and the proportion of the control class that receives a favorable outcome. Statistically significant AIR values below 0.8 can be considered \textit{prima facia} evidence of discrimination. An additional long-standing and pertinent measure of DI is SMD. SMD is often used to assess disparities in continuous features, such as income differences in employment analyses, or interest rate differences in lending. It originates from work on statistical power, and is more formally known as \textit{Cohen’s d}. SMD is equal to the difference in the average protected class outcome, $\bar{\hat{\mathbf{y}}}_p$, minus the control class outcome, $\bar{\hat{\mathbf{y}}}_c$, divided by a measure of the standard deviation of the population, $\sigma_{\hat{\mathbf{y}}}$.\footnote{There are several measures of the standard deviation of the score that are typically used: \textbf{1.} the standard deviation of the population, irrespective of protected class status, \textbf{2.} a standard deviation calculated only over the two groups being considered in a particular calculation, or \textbf{3.} a pooled standard deviation, using the standard deviations for each of the two groups with weights.} Cohen defined values of this measure to have \textit{small}, \textit{medium}, and \textit{large} effect sizes if the values exceeded 0.2, 0.5, and 0.8, respectively.

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort} to showcase the example workflow.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec:c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model fit with respect to unconstrained variants. Assuming that in high-stakes applications small fit differences on static test data do not outweigh the need for enhanced model debugging facilitated by high interpretability, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented.

@@

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the training data (e.g., LTV ratio and the no introductory rate period flag, see Figure \ref{fig:mort_xnn_glob_ridge}). This rigidity could also play a role in the performance differences between $g^\text{MGBM}(\mathbf{X})$ and $g^\text{XNN}(\mathbf{X})$ in the mortgage data not observed for the simulated data, wherein strong interactions appear to be between features with the same monotonicity constraints (e.g., $X_{\text{Friedman},1}$ and $X_{\text{Friedman},2}$, see Figure \ref{fig:sim_xnn_glob_ridge}).

PD and ICE are displayed with a histogram to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any instance with a correct schema, it's crucial to consider whether a given model learned enough about an instance to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}(\mathbf{X})$ appears to carry reasonable predictions learned from dense domains into the sparse domains of both features.

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$, and lower, exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and appear in a sparse region of property value. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}$ \cite{ice_plots}, and analysis by surrogate decision tree did show evidence of numerous potential interactions in lower predictions ranges of $g^{\text{MGBM}}(\mathbf{X})$ \cite{art_and_sci} (not shown, but available in resources discussed in Subsection \ref{ssec:soft}). Fluctuations in ICE could also be caused by overfitting or by leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.

@@

\noindent The selected individuals show an expected progression of mostly negative Shapley values at the 10\textsuperscript{th} percentile, a mixture of positive and negative Shapley values at the 50\textsuperscript{th} percentile, mostly positive Shapley values the 90\textsuperscript{th} percentile, and with globally important features driving most local model decisions. Deeper significance for Figure \ref{fig:mort_mgbm_loc} lies in the ability of Tree SHAP to accurately and consistently summarize any single $g^\text{MGBM}(\mathbf{x})$ prediction in this manner, which is generally important for enabling logical appeal or override of ML-based decisions, and is specifically important in the context of lending, where applicable regulations often require lenders to provide consumer-specific reasons for denying credit to an individual. In the US, applicable regulations are typically ECOA and FCRA, and the consumer-specific reasons are commonly known as \textit{adverse actions codes}.

@@

This text puts forward results on simulated data to provide some validation of constrained ML models, post-hoc explanation techniques, and discrimination testing methods. These same modeling, explanation, and discrimination testing approaches are then applied to more realistic mortgage data to provide an example of a responsible ML workflow for high-stakes, human-centered, or regulated ML applications. The discussed methodologies are solid steps toward interpretability, explanation, and minimal discrimination for ML decisions, which should ultimately enable increased fairness and logical appeal processes for ML decision subjects. Of course, there is more to the responsible practice of ML than interpretable models, post-hoc explanation, and discrimination testing, even from a technology perspective, and Section \ref{sec:disc} also points out numerous additional references and open source Python software assets that are available to researchers and practitioners today to increase human trust and understanding in ML systems. While the complex and messy problems of racism, sexism, privacy violations, and cyber crime can probably never be solved by technology alone, this work (and many others) illustrate numerous ways for ML practitioners to mitigate such risks.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer of, and investor in, H2O.ai and a client of BLDS, LLC. However, communications regarding XNN between Wells Fargo and the authors have been extremely limited prior to and during the drafting of this manuscript. Moreover, Wells Fargo exerted absolutely no editorial control over the text or results herein.}

@@

A finding of \textit{practical significance} means that discovered disparity is not only statistically significant, but also passes beyond a chosen threshold that would constitute \textit{prima facia} evidence of illegal discrimination.  Practical significance acknowledges that any large dataset is likely to show statistically significant differences in outcomes by class, even if those differences are not truly meaningful.  It further recognizes that there are likely to be situations where differences in outcomes are beyond a model user’s ability to correct them without significantly degrading the quality of the model. Moreover, practical significance is also needed by model builders and compliance personnel to determine whether a model should undergo remediation efforts before it is put into production. Unfortunately, guidelines for practical significance, i.e., the threshold at which any statistically significant disparity would be considered evidence of illegal discrimination, are not as frequently codified as the standards for statistical significance.  One exception, however, is in employment discrimination analyses, where the US Equal Employment Opportunity Commission (EEOC) has stated that if the AIR is below 0.80 and statistically significant, then this constitutes \textit{prima facia} evidence of discrimination, which the model user must rebut in order for the DI not to be considered illegal discrimination.\footnote{Importantly, the standard of 0.80 is not a law, but a rule of thumb for agencies tasked with enforcement of discrimination laws.  ``Adoption of Questions and Answers To Clarify and Provide a Common Interpretation of the Uniform Guidelines on Employee Selection Procedures,'' Federal Register, Volume 44, Number 43 (1979).} It is important to note that the 0.80 measure of practical significance, also known as the \textit{80\% rule} and the \textit{4/5ths rule}, is explicitly used in relation to AIR, and it is not clear that the use of this threshold is directly relevant to testing fairness for measures other than the AIR.

>>>>

% Authors, for the paper (add full first names)
\Author{Navdeep Gill $^{1, \dagger}$, Patrick Hall $^{1,3,\dagger,*}$, Kim Montgomery $^{1,\dagger}$, and Nicholas Schmidt $^{2, \dagger, *}$}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBMs) and explainable neural networks (XNNs), a deep learning architecture well-suited for structured data, are assessed on simulated data and publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and with global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents, adverse impact ratio (AIR), marginal effect (ME), and standardized mean difference (SMD), along with straightforward group fairness measures. By combining interpretable models, post-hoc explanations, and discrimination testing with accessible software tools, this text aims to provide a template workflow for important ML applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/gx/en/issues/data-and-analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}, and ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of ML opaqueness, discrimination, privacy harms, and security vulnerabilities have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, mostly in the form of interpretable models, post-hoc explanation, and DI and discrimination testing, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve Supervisory and Regulatory (SR) Letter 11-7, and the European Union (EU) General Data Protection Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

The same can be said of the selected explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. Presented explanation techniques include PD and ICE (see Subsection \ref{ssec:pd_ice}) and Shapley values (see Subsection \ref{ssec:shap}) \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discrimination testing methods discussed (see Subsection \ref{ssec:di}) include AIR, ME, and SMD \cite{feldman2015certifying}, \cite{cohen1988statistical}, \cite{cohen1992power}.\footnote{See: Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix measures are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available on GitHub (see Subsection \ref{ssec:soft}).

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}Ms), which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}. XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some instance , $\mathbf{x}$, is defined as:

@@

Because many current technical discussions of fairness in ML appear inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in regulated, high-stakes employment and financial decisions. The discussed measures are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination in US legal and regulatory settings, and Appendix \ref{a_sec:prac_v_stat} for remarks on practical vs. statistical significance for discrimination measures. One such common measure of DI used in US litigation and regulatory settings is ME. ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

\noindent where $\hat{\mathbf{y}}$ are the model decisions, $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), $p$ indicates a protected group, and $\text{Pr}(\cdot)$ is the operator for conditional probability.  ME is a favored DI measure used by the US Consumer Financial Protection Bureau (CFPB), the primary agency charged with regulating fair lending laws at the largest US lending institutions and for various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important DI measure is AIR, more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

\noindent AIR is equal to the ratio of the proportion of the protected class that receives a favorable outcome and the proportion of the control class that receives a favorable outcome. Statistically significant AIR values below 0.8 can be considered \textit{prima facie} evidence of discrimination. An additional long-standing and pertinent measure of DI is SMD. SMD is often used to assess disparities in continuous features, such as income differences in employment analyses, or interest rate differences in lending. It originates from work on statistical power, and is more formally known as \textit{Cohen’s d}. SMD is equal to the difference in the average protected class outcome, $\bar{\hat{\mathbf{y}}}_p$, minus the control class outcome, $\bar{\hat{\mathbf{y}}}_c$, divided by a measure of the standard deviation of the population, $\sigma_{\hat{\mathbf{y}}}$.\footnote{There are several measures of the standard deviation of the score that are typically used: \textbf{1.} the standard deviation of the population, irrespective of protected class status, \textbf{2.} a standard deviation calculated only over the two groups being considered in a particular calculation, or \textbf{3.} a pooled standard deviation, using the standard deviations for each of the two groups with weights.} Cohen defined values of this measure to have \textit{small}, \textit{medium}, and \textit{large} effect sizes if the values exceeded 0.2, 0.5, and 0.8, respectively.

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort} to showcase the example workflow.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec:c_v_uc_sim}, the constrained variants of both model architectures do not show large differences in model fit with respect to unconstrained variants. Assuming that in high-stakes applications small fit differences on static test data do not outweigh the need for enhanced model debugging facilitated by high interpretability, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented.

@@

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it is also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the training data (e.g., LTV ratio and the no introductory rate period flag, see Figure \ref{fig:mort_xnn_glob_ridge}). This rigidity could also play a role in the performance differences between $g^\text{MGBM}(\mathbf{X})$ and $g^\text{XNN}(\mathbf{X})$ in the mortgage data not observed for the simulated data, wherein strong interactions appear to be between features with the same monotonicity constraints (e.g., $X_{\text{Friedman},1}$ and $X_{\text{Friedman},2}$, see Figure \ref{fig:sim_xnn_glob_ridge}).

PD and ICE are displayed with a histogram to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any instance with a correct schema, it is crucial to consider whether a given model learned enough about an instance to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}(\mathbf{X})$ appears to carry reasonable predictions learned from dense domains into the sparse domains of both features.

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$ and lower exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior, while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and appear in a sparse region of property value. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}$ \cite{ice_plots}, and analysis by surrogate decision tree did show evidence of numerous potential interactions in lower predictions ranges of $g^{\text{MGBM}}(\mathbf{X})$ \cite{art_and_sci} (not shown, but available in resources discussed in Subsection \ref{ssec:soft}). Fluctuations in ICE could also be caused by overfitting or by leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.

@@

\noindent The selected individuals show an expected progression of mostly negative Shapley values at the 10\textsuperscript{th} percentile, a mixture of positive and negative Shapley values at the 50\textsuperscript{th} percentile, mostly positive Shapley values at the 90\textsuperscript{th} percentile, and with globally important features driving most local model decisions. Deeper significance for Figure \ref{fig:mort_mgbm_loc} lies in the ability of Tree SHAP to accurately and consistently summarize any single $g^\text{MGBM}(\mathbf{x})$ prediction in this manner, which is generally important for enabling logical appeal or override of ML-based decisions, and is specifically important in the context of lending, where applicable regulations often require lenders to provide consumer-specific reasons for denying credit to an individual. In the US, applicable regulations are typically ECOA and FCRA, and the consumer-specific reasons are commonly known as \textit{adverse actions codes}.

@@

This text puts forward results on simulated data to provide some validation of constrained ML models, post-hoc explanation techniques, and discrimination testing methods. These same modeling, explanation, and discrimination testing approaches are then applied to more realistic mortgage data to provide an example of a responsible ML workflow for high-stakes, human-centered, or regulated ML applications. The discussed methodologies are solid steps toward interpretability, explanation, and minimal discrimination for ML decisions, which should ultimately enable increased fairness and logical appeal processes for ML decision subjects. Of course, there is more to the responsible practice of ML than interpretable models, post-hoc explanation, and discrimination testing, even from a technology perspective, and Section \ref{sec:disc} also points out numerous additional references and open source Python software assets that are available to researchers and practitioners today to increase human trust and understanding in ML systems. While the complex and messy problems of racism, sexism, privacy violations, and cyber crime can probably never be solved by technology alone, this work and many others illustrate numerous ways for ML practitioners to mitigate such risks.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer of, and investor in, H2O.ai and a client of BLDS, LLC. However, communications regarding XNN between Wells Fargo and Patrick Hall at H2O.ai have been extremely limited prior to and during the drafting of this manuscript. Moreover, Wells Fargo exerted absolutely no editorial control over the text or results herein.}

@@

A finding of \textit{practical significance} means that discovered disparity is not only statistically significant, but also passes beyond a chosen threshold that would constitute \textit{prima facie} evidence of illegal discrimination.  Practical significance acknowledges that any large dataset is likely to show statistically significant differences in outcomes by class, even if those differences are not truly meaningful.  It further recognizes that there are likely to be situations where differences in outcomes are beyond a model user’s ability to correct them without significantly degrading the quality of the model. Moreover, practical significance is also needed by model builders and compliance personnel to determine whether a model should undergo remediation efforts before it is put into production. Unfortunately, guidelines for practical significance, i.e., the threshold at which any statistically significant disparity would be considered evidence of illegal discrimination, are not as frequently codified as the standards for statistical significance.  One exception, however, is in employment discrimination analyses, where the US Equal Employment Opportunity Commission (EEOC) has stated that if the AIR is below 0.80 and statistically significant, then this constitutes \textit{prima facie} evidence of discrimination, which the model user must rebut in order for the DI not to be considered illegal discrimination.\footnote{Importantly, the standard of 0.80 is not a law, but a rule of thumb for agencies tasked with enforcement of discrimination laws.  ``Adoption of Questions and Answers To Clarify and Provide a Common Interpretation of the Uniform Guidelines on Employee Selection Procedures,'' Federal Register, Volume 44, Number 43 (1979).} It is important to note that the 0.80 measure of practical significance, also known as the \textit{80\% rule} and the \textit{4/5ths rule}, is explicitly used in relation to AIR, and it is not clear that the use of this threshold is directly relevant to testing fairness for measures other than the AIR.
